Exact expressions for double descent and implicit regularization \\ via surrogate random design

### 2022-08-23 design/aistats/bayesian.tex
Updates from Overleaf

\newif\ifisarxiv
\isarxivfalse

% ready for submission
\ifisarxiv
  \documentclass[11pt]{article}
  \usepackage{fullpage}
\else
  \documentclass[twoside]{article}
  \usepackage[accepted]{aistats2020}

  % \special{papersize = 8.5in, 11in}
  % \setlength{\pdfpageheight}{11in}
  % \setlength{\pdfpagewidth}{8.5in}

  \usepackage{natbib}
  \renewcommand{\bibname}{References}
  \renewcommand{\bibsection}{\subsubsection*{\bibname}}
\fi

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{bm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{xfrac}
\usepackage{microtype}      % microtypography
\usepackage{color}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}

\input{../shortdefs}
\newcommand{\mub}{\bm{\mu}}

\usepackage{grffile}

@@

\ifisarxiv
  \date{}
  \def\And{\and}
\fi

\ifisarxiv
  \maketitle
\fi

\begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
\end{abstract}

@@

\bibliographystyle{apalike}
\bibliography{../pap}

\newpage
\appendix

\onecolumn

>>>>

\newif\ifisarxiv
\isarxivtrue

% ready for submission
% \ifisarxiv
%   \documentclass[11pt]{article}
%   \usepackage{fullpage}
% \else
%   \documentclass[twoside]{article}
%   \usepackage[accepted]{aistats2020}

%   % \special{papersize = 8.5in, 11in}
%   % \setlength{\pdfpageheight}{11in}
%   % \setlength{\pdfpagewidth}{8.5in}

%   \usepackage{natbib}
%   \renewcommand{\bibname}{References}
%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
% \fi
\documentclass[thesis.tex]{subfiles}

@@

% \ifisarxiv
%   \date{}
%   \def\And{\and}
% \fi

% \ifisarxiv
%   \maketitle
% \fi

% \begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
% \end{abstract}

@@

% \bibliographystyle{apalike}
% \bibliography{../pap}

% \newpage
% \appendix

% \onecolumn

### 2022-08-23 design/bayesian.tex
Commit original repos
Updates from Overleaf

% ready for submission
\ifisarxiv
  \documentclass[11pt]{article}
  \usepackage{fullpage}

  \usepackage{natbib}
\else
  \documentclass[twoside]{article}
  \usepackage{aistats2020}

  \usepackage{natbib}
  \renewcommand{\bibname}{References}
  \renewcommand{\bibsection}{\subsubsection*{\bibname}}
\fi

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsmath}
\usepackage{bm}
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{xfrac}
\usepackage{microtype}      % microtypography
\usepackage{color}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{wrapfig}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{caption}

\input{shortdefs}
\newcommand{\mub}{\bm{\mu}}

\usepackage{grffile}

\begin{document}
\ifisarxiv
  \title{Bayesian experimental design using regularized \\
    determinantal point processes}
  \author{%
          Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
  Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
   Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
    % examples of more authors
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \AND
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
    % \And
    % Coauthor \\
    % Affiliation \\
    % Address \\
    % \texttt{email} \\
  }
\else
  \twocolumn[
  \aistatstitle{Bayesian experimental design \\using regularized 
    determinantal point processes}
\aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
\aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
}
  % \aistatsauthor{%
  %   Micha{\l } Derezi\'{n}ski
  %   \And
  %   Feynman Liang
  %   \And
  %   Michael W. Mahoney
  % }
  % \aistatsaddress {
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mderezin@berkeley.edu}
  %   \And
  %   Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{feynman@berkeley.edu}
  %   \And
  %   ICSI and Department of Statistics\\
  %   University of California, Berkeley\\
  %   \texttt{mmahoney@stat.berkeley.edu}
  % } 
  ]
\fi

% The \author macro works with any number of authors. There are two commands
% used to separate the names and addresses of multiple authors: \And and \AND.
%
% Using \And between authors leaves it to LaTeX to determine where to break the
% lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% authors names on the first line, and the last on the second line, try using
% \AND instead of \And before the third author name.

\ifisarxiv
  \date{}
  \def\And{\and}
\fi

\ifisarxiv
  \maketitle
\fi

\begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
\end{abstract}

@@

\bibliographystyle{apalike}
\bibliography{pap}

>>>>

% ready for submission
% \ifisarxiv
%   \documentclass[11pt]{article}
%   \usepackage{fullpage}

%   \usepackage{natbib}
% \else
%   \documentclass[twoside]{article}
%   \usepackage{aistats2020}

%   \usepackage{natbib}
%   \renewcommand{\bibname}{References}
%   \renewcommand{\bibsection}{\subsubsection*{\bibname}}
% \fi

% \usepackage[utf8]{inputenc} % allow utf-8 input
% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
% \usepackage{amsfonts}       % blackboard math symbols
% \usepackage{amsmath}
% \usepackage{bm}
% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
% \usepackage{xfrac}
% \usepackage{microtype}      % microtypography
% \usepackage{color}
% \usepackage{graphicx}
% \usepackage{tikz}
% \usepackage{wrapfig}
% \usepackage{algorithm}
% \usepackage{algorithmic}
% \usepackage{caption}

% \input{shortdefs}
% \newcommand{\mub}{\bm{\mu}}

% \usepackage{grffile}

\documentclass[thesis.tex]{subfiles}

\begin{document}
% \ifisarxiv
%   \title{Bayesian experimental design using regularized \\
%     determinantal point processes}
%   \author{%
%           Micha{\l } Derezi\'{n}ski \\
%   Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{mderezin@berkeley.edu}\\
%   \And
%   Feynman Liang \\
%   Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{feynman@berkeley.edu}
%   \And
%   Michael W. Mahoney\\
%   ICSI and Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{mmahoney@stat.berkeley.edu}
%     % examples of more authors
%     % \And
%     % Coauthor \\
%     % Affiliation \\
%     % Address \\
%     % \texttt{email} \\
%     % \AND
%     % Coauthor \\
%     % Affiliation \\
%     % Address \\
%     % \texttt{email} \\
%     % \And
%     % Coauthor \\
%     % Affiliation \\
%     % Address \\
%     % \texttt{email} \\
%     % \And
%     % Coauthor \\
%     % Affiliation \\
%     % Address \\
%     % \texttt{email} \\
%   }
% \else
%   \twocolumn[
%   \aistatstitle{Bayesian experimental design \\using regularized 
%     determinantal point processes}
% \aistatsauthor{ Author 1 \And Author 2 \And  Author 3 }
% \aistatsaddress{ Institution 1 \And  Institution 2 \And Institution 3
% }
%   % \aistatsauthor{%
%   %   Micha{\l } Derezi\'{n}ski
%   %   \And
%   %   Feynman Liang
%   %   \And
%   %   Michael W. Mahoney
%   % }
%   % \aistatsaddress {
%   %   Department of Statistics\\
%   %   University of California, Berkeley\\
%   %   \texttt{mderezin@berkeley.edu}
%   %   \And
%   %   Department of Statistics\\
%   %   University of California, Berkeley\\
%   %   \texttt{feynman@berkeley.edu}
%   %   \And
%   %   ICSI and Department of Statistics\\
%   %   University of California, Berkeley\\
%   %   \texttt{mmahoney@stat.berkeley.edu}
%   % } 
%   ]
% \fi

% % The \author macro works with any number of authors. There are two commands
% % used to separate the names and addresses of multiple authors: \And and \AND.
% %
% % Using \And between authors leaves it to LaTeX to determine where to break the
% % lines. Using \AND forces a line break at that point. So, if LaTeX puts 3 of 4
% % authors names on the first line, and the last on the second line, try using
% % \AND instead of \And before the third author name.

% \ifisarxiv
%   \date{}
%   \def\And{\and}
% \fi

% \ifisarxiv
%   \maketitle
% \fi

% \begin{abstract}
We establish a fundamental connection between Bayesian
experimental design and determinantal point processes
(DPPs). Experimental design is a classical task in combinatorial
optimization, where we wish to select a small subset of $d$-dimensional
 vectors to minimize a statistical optimality criterion.
 We show that a new regularized variant of
 DPPs can be used to design
efficient algorithms for finding $(1+\epsilon)$-approximate solutions
to experimental design under four commonly used optimality
criteria: A-, C-, D- and V-optimality. A key novelty is that we offer
improved guarantees under the Bayesian framework.
Our algorithm returns a $(1+\epsilon)$-approximate solution when the
subset size $k$ is
$\Omega(\frac{d_{\A}}{\epsilon} + \frac{\log
1/\epsilon}{\epsilon^2})$, where $d_{\A}\ll d$ is an effective dimension
determined by prior knowledge (via a precision matrix $\A$). This is the first
approximation guarantee where the dependence on $d$ is
replaced by an effective dimension. Moreover, the time complexity
of our algorithm significantly improves on existing approaches
with comparable guarantees. 
% \end{abstract}

@@

% \bibliographystyle{apalike}
% \bibliography{pap}

### 2022-08-23 design/continuous_nips/continuous.tex
Updates from Overleaf

 \ifisarxiv
 \documentclass[11pt]{article}
 \usepackage[round]{natbib}
 \usepackage{fullpage}
 \usepackage{amsmath}
 \usepackage{amssymb}
 \usepackage{amsthm}
 \usepackage{mathabx}
 \usepackage{color}
 \usepackage{cancel}
 \usepackage{wrapfig}
 \usepackage{subfigure}
 \usepackage{caption}
 \usepackage{xfrac}
 \usepackage{algorithm}
 \usepackage{algorithmic}
 \usepackage{mdframed}
 \usepackage{hyperref}
 \usepackage{xcolor}
 \hypersetup{
 	colorlinks,
 	linkcolor={red!40!gray},
 	citecolor={blue!40!gray},
 	urlcolor={blue!70!gray}
 }
 \usepackage{cleveref}

@@

\else

\documentclass{article}
\usepackage[final]{neurips_2020}

\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography

\usepackage{amsmath}
\usepackage{cancel}
\usepackage{cleveref}
\usepackage{subfigure}
\usepackage{todonotes}
 \usepackage{wrapfig}
\usepackage{xcolor}
\usepackage{xfrac}
\hypersetup{
	colorlinks,
	linkcolor={red!40!gray},
	citecolor={blue!40!gray},
	urlcolor={blue!70!gray}
}

\title{Exact expressions for double descent and implicit regularization via
surrogate random design}
  \author{%
Micha{\l } Derezi\'{n}ski \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mderezin@berkeley.edu}\\
  \And
Feynman Liang \\
  Department of Statistics\\
  University of California, Berkeley\\
  \texttt{feynman@berkeley.edu}
  \And
Michael W. Mahoney\\
  ICSI and Department of Statistics\\
  University of California, Berkeley\\
  \texttt{mmahoney@stat.berkeley.edu}
}
% \usepackage{times}

% \coltauthor{%
%   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Zhenyu Liao} \Email{TODO}\\
%   \addr Department of Statistics\\
%   University of California, Berkeley\\
%   \AND
%   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
%   \addr ICSI and Department of Statistics\\
%   University of California, Berkeley
% }

% \usepackage{float}
% \usepackage{wrapfig}

\fi

\input{../shortdefs}

\newcommand{\RED}{\color[rgb]{0.70,0,0}}

\begin{document}
\maketitle

\begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing a special
  determinantal point process  which we call surrogate random
  design, to replace the standard i.i.d.~design of the training
  sample. This surrogate design admits exact expressions for the mean
  squared error of the estimator while preserving the key properties
  of the standard design. We also establish an exact implicit
  regularization result for over-parameterized training samples. In
  particular, we show that, for the surrogate design, the implicit bias
  of the unregularized minimum norm estimator precisely corresponds to
  solving a ridge-regularized least squares problem on the population
  distribution. In our analysis we introduce a new mathematical tool of
  independent interest: the class of random matrices for which
  determinant commutes with expectation.
\end{abstract}

@@

\begin{figure}[t]
  \centering
  \subfigure[%
    Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
    numerical estimates even for non-isotropic
    features. Eigenvalue decay leads to a steeper
    descent curve in the under-determined regime ($n<d$).]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-intro}
  }
  \hfill
  \subfigure[%
    The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.]{%
    \includegraphics[width=0.48\textwidth]{figs/descent-shrinkage}
  }
  \caption{Illustration of the main results for $d=100$ and
    $\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
    eigenvalues decaying exponentially and scaled so that
    $\tr(\Sigmab^{-1})=d$. We use our surrogate
    formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
    \ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
    lines), accompanied by the empirical estimates based on the standard
    i.i.d.~design (error bars are three times the standard error of the
    mean). We consider three different condition numbers $\kappa$ of
    $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
    i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
    $\w^*=\frac1{\sqrt{d}}\one$.}
  %\vspace{-5mm}
  \label{f:intro}
\end{figure}

@@

\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.44\fi\textwidth}
  \ifisarxiv\else \vspace{-.2cm}\fi
  \centering
  \includegraphics[width=.43\textwidth]{figs/descent-model-log}
  \ifisarxiv\else \vspace{-3mm}\fi
  \caption{
    Surrogate MSE as a function of $d/n$, with $n$
    fixed to $100$ and varying dimension $d$ and condition number
    $\kappa$, for signal-to-noise ratio $\mathrm{SNR}=1$.}
  \label{f:model}
  \ifisarxiv\else  \vspace{-.5cm}\fi
\end{wrapfigure}

@@

\ifisarxiv
  \bibliographystyle{plainnat}
\else
  \bibliographystyle{alpha}
\fi

\bibliography{../pap}

\ifisarxiv\else\newpage\fi

\appendix

>>>>

%  \ifisarxiv
 \documentclass[thesis.tex]{subfiles}
%  \usepackage[round]{natbib}
%  \usepackage{fullpage}
%  \usepackage{amsmath}
%  \usepackage{amssymb}
%  \usepackage{amsthm}
%  \usepackage{mathabx}
%  \usepackage{color}
%  \usepackage{cancel}
%  \usepackage{wrapfig}
%  \usepackage{subfigure}
%  \usepackage{caption}
%  \usepackage{xfrac}
%  \usepackage{algorithm}
%  \usepackage{algorithmic}
%  \usepackage{mdframed}
%  \usepackage{hyperref}
%  \usepackage{xcolor}
%  \hypersetup{
%  	colorlinks,
%  	linkcolor={red!40!gray},
%  	citecolor={blue!40!gray},
%  	urlcolor={blue!70!gray}
%  }
%  \usepackage{cleveref}

@@

% \else

% \documentclass{article}
% \usepackage[final]{neurips_2020}

% \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
% \usepackage{hyperref}       % hyperlinks
% \usepackage{url}            % simple URL typesetting
% \usepackage{booktabs}       % professional-quality tables
% \usepackage{amsfonts}       % blackboard math symbols
% \usepackage{nicefrac}       % compact symbols for 1/2, etc.
% \usepackage{microtype}      % microtypography

% \usepackage{amsmath}
% \usepackage{cancel}
% \usepackage{cleveref}
% \usepackage{subfigure}
% \usepackage{todonotes}
%  \usepackage{wrapfig}
% \usepackage{xcolor}
% \usepackage{xfrac}
% \hypersetup{
% 	colorlinks,
% 	linkcolor={red!40!gray},
% 	citecolor={blue!40!gray},
% 	urlcolor={blue!70!gray}
% }

% \title{Exact expressions for double descent and implicit regularization via
% surrogate random design}
%   \author{%
% Micha{\l } Derezi\'{n}ski \\
%   Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{mderezin@berkeley.edu}\\
%   \And
% Feynman Liang \\
%   Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{feynman@berkeley.edu}
%   \And
% Michael W. Mahoney\\
%   ICSI and Department of Statistics\\
%   University of California, Berkeley\\
%   \texttt{mmahoney@stat.berkeley.edu}
% }
% % \usepackage{times}

% % \coltauthor{%
% %   \Name{Micha{\l } Derezi\'{n}ski} \Email{mderezin@berkeley.edu}\\
% %   \addr Department of Statistics\\
% %   University of California, Berkeley\\
% %   \AND
% %   \Name{Feynman Liang} \Email{feynman@berkeley.edu}\\
% %   \addr Department of Statistics\\
% %   University of California, Berkeley\\
% %   \AND
% %   \Name{Zhenyu Liao} \Email{TODO}\\
% %   \addr Department of Statistics\\
% %   University of California, Berkeley\\
% %   \AND
% %   \Name{Michael W. Mahoney} \Email{mmahoney@stat.berkeley.edu}\\
% %   \addr ICSI and Department of Statistics\\
% %   University of California, Berkeley
% % }

% % \usepackage{float}
% % \usepackage{wrapfig}

% \fi

% \input{../shortdefs}

% \newcommand{\RED}{\color[rgb]{0.70,0,0}}

\begin{document}
% \maketitle

% \begin{abstract}
  Double descent refers to the phase transition that is exhibited by
  the generalization error of unregularized learning models when varying the ratio
  between the number of parameters and the number of training
  samples. The recent success of highly over-parameterized machine learning
  models such as deep neural networks has motivated a theoretical analysis of
  the double descent phenomenon in classical models such as linear
  regression which can also generalize well in the over-parameterized
  regime. We provide the first exact non-asymptotic
  expressions for double descent of the minimum norm linear
  estimator. Our approach involves constructing a special
  determinantal point process  which we call surrogate random
  design, to replace the standard i.i.d.~design of the training
  sample. This surrogate design admits exact expressions for the mean
  squared error of the estimator while preserving the key properties
  of the standard design. We also establish an exact implicit
  regularization result for over-parameterized training samples. In
  particular, we show that, for the surrogate design, the implicit bias
  of the unregularized minimum norm estimator precisely corresponds to
  solving a ridge-regularized least squares problem on the population
  distribution. In our analysis we introduce a new mathematical tool of
  independent interest: the class of random matrices for which
  determinant commutes with expectation.
% \end{abstract}

@@

\begin{figure}[t]
  \centering
  \begin{subfigure}{0.49\textwidth}%
    \subcaption{Surrogate MSE expressions (Theorem \ref{t:mse}) closely match
    numerical estimates even for non-isotropic
    features. Eigenvalue decay leads to a steeper
    descent curve in the under-determined regime ($n<d$).}
    \includegraphics[width=0.48\textwidth]{../figs/descent-intro}
  \end{subfigure}
  \hfill
  \begin{subfigure}{0.49\textwidth}
    \subcaption{The mean of the estimator $\X^\dagger\y$ exhibits
    shrinkage which closely matches the shrinkage of a
    ridge-regularized least squares optimum (theory lines), as characterized by
    Theorem \ref{t:unbiased}.}
    \includegraphics[width=0.48\textwidth]{../figs/descent-shrinkage}
  \end{subfigure}
  \caption{Illustration of the main results for $d=100$ and
    $\mu=\Nc(\zero,\Sigmab)$ where $\Sigmab$ is diagonal with
    eigenvalues decaying exponentially and scaled so that
    $\tr(\Sigmab^{-1})=d$. We use our surrogate
    formulas to plot (a) the MSE (Theorem \ref{t:mse}) and (b) the norm of the expectation (Theorem
    \ref{t:unbiased}) of the Moore-Penrose estimator (\emph{theory}
    lines), accompanied by the empirical estimates based on the standard
    i.i.d.~design (error bars are three times the standard error of the
    mean). We consider three different condition numbers $\kappa$ of
    $\Sigmab$, with \emph{isotropic} corresponding to $\kappa=1$,
    i.e., $\Sigmab=\I$. We use $\sigma^2=1$ and
    $\w^*=\frac1{\sqrt{d}}\one$.}
  %\vspace{-5mm}
  \label{f:intro}
\end{figure}

@@

\begin{wrapfigure}{r}{\ifisarxiv 0.5\else 0.44\fi\textwidth}
  \ifisarxiv\else \vspace{-.2cm}\fi
  \centering
  \includegraphics[width=.43\textwidth]{../figs/descent-model-log}
  \ifisarxiv\else \vspace{-3mm}\fi
  \caption{
    Surrogate MSE as a function of $d/n$, with $n$
    fixed to $100$ and varying dimension $d$ and condition number
    $\kappa$, for signal-to-noise ratio $\mathrm{SNR}=1$.}
  \label{f:model}
  \ifisarxiv\else  \vspace{-.5cm}\fi
\end{wrapfigure}

@@

% \ifisarxiv
%   \bibliographystyle{plainnat}
% \else
%   \bibliographystyle{alpha}
% \fi

% \bibliography{../pap}

% \ifisarxiv\else\newpage\fi

% \appendix
